---
title: "The Causal Impacts of 'Clutch' Situations on NBA Shooting Percentages"
author: "Alexander Hinton"
always_allow_html: true
output:
 github_document:
    pandoc_args: --webtex
    toc: true
    toc_depth: 1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(Matching)
library(knitr)
library(kableExtra)
library(ggrepel)
library(gridExtra)
```

### Background

Based on the perceived ability to perform in the most important situations, fans, media members and players alike love to use the term `clutch` to describe certain players. While highlight reels and biased memories may make it seem like our favourite players always deliver when the game is on the line, the reality is much different. A recent [article by Seth Partnow](https://theathletic.com/1142773/2019/08/20/the-clutch-factor-analyzing-hero-shots-in-the-nba/) showed that the average shooting percentage in `clutch` situations was only around $30\%$, much lower than the conversion rate across all shot types. Even players widely considered to be "clutch", such as Kobe Bryant, had clutch shooting percentages right around this $30\%$ mark. The natural follow up question which arises is what makes these shots have such a low conversion rate. There are two main drivers which I hypothesize: 

- Shots taken in these clutch situations are simply very difficult shots on average, in terms of defensive pressure and distance from the basket. 
- The intensity of the situation, with the crowd standing on their feet, and the increased adrenaline causing faster heartrates and sweaty palms causes players to be less accurate in their shooting.
\newline
Of course, a combination of these two effects is also possible.<br>

I aim to use matching techniques to perform a causal inference based analysis to determine whether there is in fact a causal impact of clutch situations reducing shooting percentage in the NBA, beyond simply taking more difficult shots. 


#### Defining clutch
Before continuing, the definition of what constitutes a clutch shot in this analysis needs to be established, as this term is inherently subject. I delegated any shot in the last $30$" of a game, where the score was within $3$ points (a one possession game) as a clutch shot, which is similar to the definition used by Partnow in the article referenced above. These are the shots where a timeout will often be taken beforehand, the crowd is all on their feet, and the nerves would be at their peak. 

In terms of causal inference lingo, the `treatment` group will be these shot attempts which were taken in the clutch situations, while the `control` group will be all shot attempts outside of these situations, where the stakes and pressure of the moment are significantly less on average. 

<!-- In order to isolate potential causal impacts I will use propensity score matching, as well as a more handpicked case-control matching strategy to find pairs of shots deemed to be very similar, with the difference being whether each shot was in a clutch situation or not. -->


#### Gathering data
In order to estimate a potential causal impact of the clutch situation on shooting accuracy, the difficulty of the shot must be controlled for. This requires quite granular data about each shot attempt. Fortunately, the website [NBA-savant](https://www.nbasavant.com/) provides detailed data on all NBA shots, including shot distance, distance to the closest defender, name of the closest defender, number of dribbles before the shot, etc. Unfortunately, the NBA stopped making this data public in January 2016, meaning the most recent NBA seasons could not be used in this analysis. I used all regular season shots from the 2013-2014 season, 2014-2015 season, and the 2015-2016 season until the granular shot data stopped being released. <br>

While game clock information was available from the shot data, the score at the time the shot was taken was not. Using my definition of clutch from above, this score information is required. This info was gathered by merging the shot data with play-by-play data, and adding in the current score information which was unavailable in the shot charts. 

<!-- Approximately $80\%$ of shots were able to matched with shot events in the play by play data (small discrepancies in game clock at the time of shot meant not all shots were matched).  -->

Once I had gathered the shot information I separated shots into four categories: clutch and non-clutch twos, and clutch and non-clutch threes. For two pointers, I only included shots that were from at least $10$ feet, as I wanted to be comparing jumpers, and did not think the same clutch effect would apply to lay-ups, dunks, put-backs etc. The control group of non-clutch shots consisted of any shot taken in the first three quarters of a game. I considered lumping all shot types together in order to increase sample sizes, but thought that if there truly is a crunch time, or clutch effect on shooting accuracy it may not be the same between two pointers and three pointers, and they should be treated separately. 
\newline 
Before diving into the causal inference approaches, it's important to have an idea what the data looks like.

#### Summary stats
```{r summary stats, echo = FALSE}
all_clutch_twos <- read_csv("../data/all_clutch_twos.csv", col_types = cols())
non_clutch_twos <- read_csv("../data/non_clutch_twos.csv", col_types = cols())
rbind(all_clutch_twos, non_clutch_twos) %>% 
### Summary stats 
                group_by(isClutch) %>%
                  summarize(fg_percentage = round(100*mean(shot_made_flag),1),
                 defender_height = round(mean(defender_height),1),
                 dribbles = round(mean(dribbles),1),
                 touch_time = round(mean(touch_time),1),
                 shot_dist = round(mean(shot_distance),1),
                 defend_dist = round(mean(defender_distance),1),
                 n = n()) %>%
    rename("FG percentage" = fg_percentage,
           "Defender Height"= defender_height,
           "Dribbles" = dribbles,
           "Touch time" = touch_time,
           "Shot Distance" = shot_dist,
           "Defender Distance" = defend_dist,
           "Count" = n) %>% 
 kable(caption = "Summary Stats: Averages for Clutch and Non-Clutch Two Point Shots",
      format = 'html') %>%
  kable_styling()
```

```{r summary stats 3s, echo = FALSE}
all_clutch_threes <- read_csv("../data/all_clutch_threes.csv", col_types = cols())
non_clutch_threes <- read_csv("../data/non_clutch_threes.csv", col_types = cols())
rbind(all_clutch_threes, non_clutch_threes) %>% 
### Summary stats 
                group_by(isClutch) %>%
                  summarize(fg_percentage = round(100*mean(shot_made_flag),1),
                 defender_height = round(mean(defender_height),1),
                 dribbles = round(mean(dribbles),1),
                 touch_time = round(mean(touch_time),1),
                 shot_dist = round(mean(shot_distance),1),
                 defend_dist = round(mean(defender_distance),1),
                 n = n()) %>%
    rename("FG percentage" = fg_percentage,
           "Defender Height"= defender_height,
           "Dribbles" = dribbles,
           "Touch time" = touch_time,
           "Shot Distance" = shot_dist,
           "Defender Distance" = defend_dist,
           "Count" = n) %>%
   kable(caption = "Summary Stats: Averages for Clutch and Non-Clutch Three Point Shots",
        format = 'html') %>%
  kable_styling()
```

Clearly there are some pretty stark differences in shooting percentages between non-clutch and clutch shots for both two pointers and three pointers. However, we are clearly not comparing apples to apples in terms of shot difficulty. For both two pointers and three pointers, the average clutch shot attempt is from farther away, has a closer defender and is off the back of a longer individual effort (more dribbles and a longer touch time). These in total make it obvious that average clutch shot attempts on average are much harder than non clutch attempts, and cannot be compared directly. A strategy to account for the difficulty of a shot is needed in order to make a fair comparison between a clutch shot and a non-clutch shot when trying to estimate a causal impact of the clutch situation on shooting accuracy. 

#### Causal Inference in Observational Studies
In an ideal world, randomized control trials are used to estimate the causal impact of a treament. In these situations, the treatment of interest is randomly allocated to the observational units (often people) in the study. Some outcome measure is calculated for both the treatment and control groups, and the difference in this outcome between groups is used as an estimate of the efficacy of the treatment. Due to the randomized nature of treatment assignment, the chance of biased estimates due to confounding is minimized. 
\newline
In situations where a randomized controlled trial is possible, observational studies can sometimes be used to estimate causal treatment effects. The main drawback and danger when using observational studies for causal inference is the potential for biased estimates due to confounding. In an observational study, treatment assignments have not been done randomly, which can be a problem if there are variables associated with both the treatment assignment and the outcome variable, which are known as confounders.  For example, in a study looking at the effectiveness of eating vegetables on health outcomes it would be naive to simply compare the outcomes of individuals eating veggie rich diets versus those who are not, as the individuals eating more veggies may also be excercising more, maybe have higher incomes, and these would both be correlated with higher health outcomes. This potential bias in estimates due to confounding variables is the main difficulty of performing causal inference on observational studies where treatments have not been explicitly assigned. 

#### Matching strategies to estimate causal impacts
Matching strategies are often used in an attempt to mimic a randomized experiment when using observational data. In matching studies, observations which received the treatment are carefully matched with observations in the control group such that the treatment and control group have similar covariate distributions.  

In this analysis the treatment group (the clutch shots), clearly have different levels for many of the covariates of interest than the control groups (the non clutch shots). As seen above, defender distance, number of dribbles, touch time, etc. are all different between the treatment group levels at least in their mean levels.

I attempted two different matching strategies in order to estimate a causal effect, the first was an exact matching approach, and the second was matching using propensity scores. In both cases I attempted to match each clutch shot to a non-clutch shot, which means the effect I am looking for is known is called the Average Treatment effect on the Treated (ATT). 

In the exact matching, I considered the following covariates to be important when quantifying the difficulty of a shot:  shot distance, defender distance, defender, number of dribbles, and touch time. 
For each clutch shot, I attempted to find a shot in the non-clutch dataframe which would match on:

- `name`: must be the same shooter
- `season`: must be in the same season
- `shot distance`: +/- 1 foot
- `defender distance`, +/- 0.5 feet
- `number of dribbles`, +/- 1
- `touch time`, +/- 0.5 seconds
- `defender height`, +/- 1 inch (Note: matching explicitly on who was the defender made the results very sparse, so I instead calculated the height of the closest defender, and matched on defender height. While not exact, I think matching on the distance to the closest defender plus the height of the defender should capture the degree the shot was contested pretty well.
- `shot type`: In the play-by-play data, a shot can be classified as a "jump shot", "fadeaway jump shot" etc., so I also matched on shot-type to control for difficulty.

All told, of the 484 clutch three point shot attempts identified and 408 clutch two point attempts, 170 of the threes, but only 70 of the two point attempts could be matched to a non-clutch shot attempt using this rigid matching criteria. 

_Note_: if more than 1 non-clutch shot matched, I would randomly select one as the matching shot.

#### Summary stats - exact matching
```{r summary stats brute force matched 2s, echo = FALSE}
# dude, re-do analysis and include shot distance!
matched_twos <- read_csv("../data/all_twos_matched.csv", col_types = cols())

### Summary stats 
matched_twos %>%
                  # include number of shots
                  group_by(isClutch) %>%
                  summarize(fg_percentage = round(100*mean(shot_made_flag),1),
                 defender_height = round(mean(defender_height),1),
                 dribbles = round(mean(dribbles),1),
                 touch_time = round(mean(touch_time),1),
                 shot_dist = round(mean(shot_distance),1),
                 defend_dist = round(mean(defender_distance),1),
                  n = n()) %>%
    rename("FG percentage" = fg_percentage,
           "Defender Height"= defender_height,
           "Dribbles" = dribbles,
           "Touch time" = touch_time,
           "Shot Distance" = shot_dist,
           "Defender Distance" = defend_dist,
           "Count" = n) %>%
   kable(caption = "Matched 2 pointers",
        format = 'html') %>%
  kable_styling()
```


```{r summary stats brute force matched 3s, echo = FALSE}
# dude, re-do analysis and include shot distance!
matched_threes <- read_csv("../data/all_threes_matched.csv", col_types = cols())

### Summary stats 
matched_threes %>%
                  # include number of shots
                  group_by(isClutch) %>%
                  summarize(fg_percentage = round(100*mean(shot_made_flag),1),
                 defender_height = round(mean(defender_height),1),
                 dribbles = round(mean(dribbles),1),
                 touch_time = round(mean(touch_time),1),
                 shot_dist = round(mean(shot_distance),1),
                 defend_dist = round(mean(defender_distance),1),
                  n = n()) %>%
    rename("FG percentage" = fg_percentage,
           "Defender Height"= defender_height,
           "Dribbles" = dribbles,
           "Touch time" = touch_time,
           "Shot Distance" = shot_dist,
           "Defender Distance" = defend_dist,
           "Count" = n) %>%
   kable(caption = "Matched 3 pointers",
        format = 'html') %>%
  kable_styling()
```

We can see that the exact matching procedure has done it's job, as the average levels of the covariates are almost identical between clutch and non clutch shots. What is interesting is that even after controlling for shot difficulty based on the observed covariates, the shooting percentages of the clutch shots are about $8-15\%$ lower than the non-clutch shot attempts.  The next step is to estimate a logistic regression model to see if these differences are statistically significant. Note that matching and controlling for covariates (another approach to isolate causal effects) do not have to be independent methods, but work best in tandem. So in the logistic regression model, the dependent variable is whether the shot was made, while the independent variables are the indicator variable of whether a shot was clutch or not (our Treatment) plus all the other control variables. 

#### Logistic regression model results
```{r echo= FALSE}
lr_twos <- glm(shot_made_flag ~ isClutch + shot_distance + defender_height + dribbles + touch_time + shot_distance + defender_distance, 
               family = "binomial", data = matched_twos)

broom::tidy(lr_twos) %>%
  mutate_if(is.numeric, round, 3) %>%
  kable(caption = "Model Parameters for two point shots",
        format = 'html') %>%
   row_spec(2, bold = TRUE) %>%
  kable_styling()
```

```{r echo= FALSE}
lr_threes <- glm(shot_made_flag ~ isClutch + shot_distance + defender_height + dribbles + touch_time + shot_distance + defender_distance, 
               family = "binomial", data = matched_threes)

broom::tidy(lr_threes) %>%
  mutate_if(is.numeric, round, 3) %>%
  kable(caption = "Model Parameters for three point shots",
        format = 'html') %>%
   row_spec(2, bold = TRUE) %>%
  kable_styling()
```
As expected based on the summary stats of the matched data, the estimate associated with the `isClutch` parameter is negative. Clutch shot attempts are estimated to have odds of success only $0.52$ (twos) or $0.69$ (threes) times the odds of success for their corresponding non-clutch shot attempts, holding all other covariates constant. Interestingly and a bit surprisingly, while there was a large difference in field goal percentage between clutch and non-clutch shot attempts even in the matched dataframe, the effect is not statistically significant at the $5\%$ significance level for both two point shots and three point shots, which is perhaps due to the reduced sample sizes. <br>


### Propensity score matching
Exact matching approaches like the one used above struggle with situations where there are many observed covariates as it becomes very difficult to find exact matches for each treated unit, and so sample sizes are greatly reduced. This difficulty has spurred on the use of what are known as propensity scores for matching purposes. 
In propensity score matching, the matching is done via the use of a statistical modelling instead of the rule based matching done above. The first step in the process is to perform logistic regression (or any other binary classifier which can output probabilities), with the treatment level (`isClutch`) as the dependent variable, and the covariates as the independent variables. The outputs of this regression are the estimated probabilitities that each observation is a clutch shot. After this estimation, we can then match each clutch shot to a non-clutch, not explicitly on each covariate value as was done above, but on the probability that each shot was a clutch shot, given it's covariate levels. In this way the distribution of covariates should be very similar between the matched clutch and non clutch groups.

#### Fit the propensity score model
Before fitting the model, we need to filter down our non clutch shots, on account of an important notion in causal inference known as positivity. Basically, what `Positivity` means is that we only look for causal impacts on observations/individuals who potentially could have received either treatment value - which in our case means players who take both clutch and non clutch shots. I can think of many players who are very unlikely to ever take a clutch shot, as these are almost always taken by the best players on a team. In light of this, I only consider non-clutch shots from players who have also taken at least one clutch shot in the propensity score modelling. This process removes about one third of all the non-clutch shots.

\newline
The covariates I will consider in the model will be the same ones as were considered for the matching experiment above:

- `Name` of shooter
- `Defender height`
- `shot distance`
- `defender distance`
- `dribbles`
- `touch time`

#### Summary stats - with propensity scores
```{r echo=FALSE}
all_pos_threes <- read_csv("../data/all_positive_threes.csv", col_types = cols())
all_pos_twos <- read_csv("../data/all_positive_twos.csv", col_types = cols())

# Fit prop score model for the threes
prop_lr_threes <- glm(isClutch ~ defender_height + shot_distance + name + defender_distance + dribbles + touch_time,
              family = 'binomial', data = all_pos_threes)
all_pos_threes['propensity_score'] <- predict.glm(prop_lr_threes,
                                                  all_pos_threes,
                                                  type = "response")
# Fit prop score model for the twos
prop_lr_twos <- glm(isClutch ~ defender_height + shot_distance + name + defender_distance + dribbles + touch_time,
              family = 'binomial', data = all_pos_twos)
all_pos_twos['propensity_score'] <- predict.glm(prop_lr_twos,
                                                  all_pos_twos,
                                                  type = "response")
```

```{r, echo=FALSE}
### Summary stats 
all_pos_threes %>%
                  # include number of shots
                  group_by(isClutch) %>%
                  summarize(fg_percentage = round(100*mean(shot_made_flag),1),
                 defender_height = round(mean(defender_height),1),
                 dribbles = round(mean(dribbles),1),
                 touch_time = round(mean(touch_time),1),
                 shot_dist = round(mean(shot_distance),1),
                 defend_dist = round(mean(defender_distance),1),
                 propensity_score = round(mean(propensity_score), 4),
                  n = n()) %>%
  rename("FG percentage" = fg_percentage,
           "Defender Height"= defender_height,
           "Dribbles" = dribbles,
           "Touch time" = touch_time,
           "Shot Distance" = shot_dist,
           "Defender Distance" = defend_dist,
           "Count" = n) %>% 
   kable(caption = "Summary Stats for 'Positive' Threes",
        format = 'html') %>%
  kable_styling()
```
```{r, echo=FALSE}
all_pos_twos %>%
                  # include number of shots
                  group_by(isClutch) %>%
                  summarize(fg_percentage = round(100*mean(shot_made_flag),1),
                 defender_height = round(mean(defender_height),1),
                 dribbles = round(mean(dribbles),1),
                 touch_time = round(mean(touch_time),1),
                 shot_dist = round(mean(shot_distance),1),
                 defend_dist = round(mean(defender_distance),1),
                 propensity_score = round(mean(propensity_score), 4),
                  n = n()) %>%
   kable(caption = "Summary Stats for 'Positive' Twos",
        format = 'html') %>%
  kable_styling()
```
  

The average propensity score for the clutch shots was more than three times higherfor clutch three point shots, and about four times higher for clutch two point shots than the average propensity scores for the non clutch twos. This basically just means the distributions of the clutch and non-clutch shots are different in terms of the covariates, which we already knew. The reason the propensity scores (or predicted probabilities of being a clutch shot) are so low is because there are so many more non-clutch shots in the sample than clutch shots, but the important thing is the large relative difference between the two groups. Next we will match observations in the treated group with the closest observation in the control group by looking at propensity scores. 

```{r, echo=FALSE}
threes_prop <- all_pos_threes  %>%  group_by(isClutch) %>%
                      sample_n(474)     %>% 
                      mutate(clutch_shots = replace(isClutch, 1, "Clutch"),
                            clutch_shots = replace(isClutch, 0, "not clutch")) %>%
                    ggplot() + aes(x = clutch_shots, y = propensity_score) +
                               geom_jitter(size = 0.4, width = 0.3, alpha = 0.5) + 
                              labs(x = "isClutch", y = "Propensity Score", 
                                   title = "Three Pointers Propensity Scores") +
                              coord_cartesian(ylim = c(0, 0.20)) 
twos_prop <- all_pos_twos %>%  group_by(isClutch) %>%
                      sample_n(408)     %>% 
                      mutate(clutch_shots = replace(isClutch, 1, "Clutch"),
                            clutch_shots = replace(isClutch, 0, "not clutch")) %>%
                    ggplot() + aes(x = clutch_shots, y = propensity_score) +
                               geom_jitter(size = 0.4, width = 0.3, alpha = 0.5) + 
                              labs(x = "isClutch", y = "Propensity Score", 
                                   title = "Two Pointers Propensity Scores") +
                              coord_cartesian(ylim = c(0, 0.20)) 

grid.arrange(twos_prop, threes_prop, nrow = 1)
```

- Note that this plot has only a random sample of the non-clutch shots for visualization purposes.


Looking at the scatter plots of propensity scores for clutch and non-clutch shots, it's clear that almost all the non clutch shots had propensity scores close to $0$, while there was a much wider distribution of propensity scores for the clutch shots. The matching process now is quite intuitive, for each shot in the clutch subset, we will find the shot in the non-clutch subset that is closest in terms of propensity scores. This does not mean the exact levels of each covariate are the same like in the brute force matching above, but it means that the probability that two shots were assigned the treatment is the same, conditional on their covariate levels. Some clutch shots appear to have propensity scores which are well above all of the non-clutch shots, so we will set a minimum threshold to determine how close two different observations need to be for matching purposes.

##### Propensity matching
```{r, echo=FALSE}
## prop score matching
ps_matching_threes <- Match(Y=all_pos_threes$propensity_score, 
                Tr=all_pos_threes$isClutch,  
                replace=TRUE, caliper=0.05,
                X=all_pos_threes$propensity_score, 
                ties = FALSE, M = 1, 
                estimand="ATT")


### Created matched df
matched_nonclutch_threes <- all_pos_threes[ps_matching_threes$index.control,]
matched_clutch_threes <- all_pos_threes[ps_matching_threes$index.treated,]
matched_ps_threes <- rbind(matched_nonclutch_threes, matched_clutch_threes)


## prop score matching twos
ps_matching_twos <- Match(Y=all_pos_twos$propensity_score,
                Tr=all_pos_twos$isClutch,
                replace=TRUE, caliper=0.05,
                X=all_pos_twos$propensity_score,
                ties = FALSE, M = 1,
                estimand="ATT")


### Created matched df
matched_nonclutch_twos <- all_pos_twos[ps_matching_twos$index.control,]
matched_clutch_twos <- all_pos_twos[ps_matching_twos$index.treated,]
matched_ps_twos <- rbind(matched_nonclutch_twos, matched_clutch_twos)
```

After matching based on propensity scores but before moving on to the final modelling stage, we need to do some inspection of the matched sets to insure that the procedure has made the control and treatment groups sufficiently similar in their covariate levels. A more formal check is done below based on the standardized bias of each covariate, but the first thing I like to do is a manual check.


```{r echo=FALSE}
### Summary stats 
matched_ps_twos %>%
                  # include number of shots
                  group_by(isClutch) %>%
                  summarize(fg_percentage = round(100*mean(shot_made_flag),1),
                 defender_height = round(mean(defender_height),1),
                 dribbles = round(mean(dribbles),1),
                 touch_time = round(mean(touch_time),1),
                 shot_dist = round(mean(shot_distance),1),
                 defend_dist = round(mean(defender_distance),1),
                 propensity_score = round(mean(propensity_score), 4),
                  n = n()) %>%
                rename("FG percentage" = fg_percentage,
                       "Defender Height"= defender_height,
                       "Dribbles" = dribbles,
                       "Touch time" = touch_time,
                       "Shot Distance" = shot_dist,
                       "Defender Distance" = defend_dist,
                       "Count" = n) %>%
   kable(caption = "Summary Stats of Two Pointers After Propensity Score Matching",
        format = 'html') %>%
  kable_styling()
```


```{r, echo=FALSE}
### Summary stats 
matched_ps_threes %>%
                  # include number of shots
                  group_by(isClutch) %>%
                  summarize(fg_percentage = round(100*mean(shot_made_flag),1),
                 defender_height = round(mean(defender_height),1),
                 dribbles = round(mean(dribbles),1),
                 touch_time = round(mean(touch_time),1),
                 shot_dist = round(mean(shot_distance),1),
                 defend_dist = round(mean(defender_distance),1),
                 propensity_score = round(mean(propensity_score), 4),
                  n = n()) %>%
                rename("FG percentage" = fg_percentage,
                       "Defender Height"= defender_height,
                       "Dribbles" = dribbles,
                       "Touch time" = touch_time,
                       "Shot Distance" = shot_dist,
                       "Defender Distance" = defend_dist,
                       "Count" = n) %>%
   kable(caption = "Summary Stats of Three Pointers After Propensity Score Matching",
        format = 'html') %>%
  kable_styling()
```



As we have not matched explicitly on each covariate, we can see some differences in some mean covariate levels between the two groups. We also see the massive benefit of using this procedure in that sample sizes can now be much larger, relative to the exact matching procedure above. Let's now check the standardized bias of each covariate, which is a ratio of the differences in mean covariate levels between the treatment and control group, divided by the standard deviation of the treatment group. 
$\frac{\bar{X_t}-\bar{X_c}}{\sigma_t}$. The maximum standardized bias for any covariate should be nor more than $0.25$ to ensure that the matched groups are similar enough to perform valid inference.

```{r, echo=FALSE}
standardized_bias <- function(col, data){
  clutch <- data %>% filter(isClutch == 1)
  non_clutch <- data %>% filter(isClutch == 0)
  
  mean_clutch <- mean(clutch %>% dplyr::select({{col}}) %>% 
                           pull())
  mean_non_clutch <- mean(non_clutch %>% dplyr::select({{col}}) %>% 
                            pull())
  sd_clutch <- sd(clutch %>% dplyr::select({{col}}) %>% 
                            pull())
  
  abs(mean_clutch - mean_non_clutch)/sd_clutch
}

# Twos
## Pre matching SB
def_height <- standardized_bias("defender_height", all_pos_twos)
dribbles <- standardized_bias("dribbles", all_pos_twos)
touch <- standardized_bias("touch_time", all_pos_twos)
shot_dist  <- standardized_bias("shot_distance", all_pos_twos)
defend_dist  <- standardized_bias("defender_distance", all_pos_twos)
pre_match_twos <- tibble(variable = c("defender_height", "dribbles", "touch time", "shot distance", "defender distance"), 
                            standardized_bias = c(def_height, dribbles, touch, shot_dist, defend_dist))

## Post matching SB
def_height <- standardized_bias("defender_height", matched_ps_twos)
dribbles <- standardized_bias("dribbles", matched_ps_twos )
touch <- standardized_bias("touch_time", matched_ps_twos)
shot_dist  <- standardized_bias("shot_distance", matched_ps_twos)
defend_dist  <- standardized_bias("defender_distance", matched_ps_twos)
post_match_twos <- tibble(variable = c("defender_height", "dribbles", "touch time", "shot distance", "defender distance"), 
                            standardized_bias = c(def_height, dribbles, touch, shot_dist, defend_dist))
# Threes
## Pre matching SB
def_height <- standardized_bias("defender_height", all_pos_threes)
dribbles <- standardized_bias("dribbles", all_pos_threes)
touch <- standardized_bias("touch_time", all_pos_threes)
shot_dist  <- standardized_bias("shot_distance", all_pos_threes)
defend_dist  <- standardized_bias("defender_distance", all_pos_threes)
pre_match_threes <- tibble(variable = c("defender_height", "dribbles", "touch time", "shot distance", "defender distance"), 
                            standardized_bias = c(def_height, dribbles, touch, shot_dist, defend_dist))

## Post matching SB
def_height <- standardized_bias("defender_height", matched_ps_threes)
dribbles <- standardized_bias("dribbles", matched_ps_threes )
touch <- standardized_bias("touch_time", matched_ps_threes)
shot_dist  <- standardized_bias("shot_distance", matched_ps_threes)
defend_dist  <- standardized_bias("defender_distance", matched_ps_threes)
post_match_threes <- tibble(variable = c("defender_height", "dribbles", "touch time", "shot distance", "defender distance"), 
                            standardized_bias = c(def_height, dribbles, touch, shot_dist, defend_dist))
  
# threes
bias_pre_threes <- mutate(pre_match_threes, Type = 0)
bias_post_threes <- mutate(post_match_threes, Type = 1)
bias_threes <- rbind(bias_pre_threes, bias_post_threes)

# two
bias_pre_twos <- mutate(pre_match_twos, Type = 0)
bias_post_twos <- mutate(post_match_twos, Type = 1)
bias_twos <- rbind(bias_pre_twos, bias_post_twos)
```


```{r, echo=FALSE}
p3 <- ggplot(bias_threes, aes(x = Type, y =standardized_bias, group = variable)) + 
  geom_point() + 
  geom_line() +
  theme(legend.position = "none") +
  ggtitle("Pre/post matched - Threes") + 
  xlab("") + ylab("Standardized Bias")  + 
  scale_x_continuous(breaks = c(0, 1), 
                     labels = c("Pre-matched", "Post-matched"), lim = c(-0.05, 1.05)) + 
  geom_label_repel(data=filter(bias_threes, Type == 0, standardized_bias > 0.2),
                   aes(label=variable)) +
  geom_hline(yintercept = 0.25, lty = 2, colour = "red")

p2 <- ggplot(bias_twos, aes(x = Type, y =standardized_bias, group = variable)) + 
  geom_point() + 
  geom_line() +
  theme(legend.position = "none") +
  ggtitle("Pre/post matched - Twos") + 
  xlab("") + ylab("Standardized Bias")  + 
  scale_x_continuous(breaks = c(0, 1), 
                     labels = c("Pre-matched", "Post-matched"), lim = c(-0.05, 1.05)) + 
  geom_label_repel(data=filter(bias_twos, Type == 0, standardized_bias > 0.2),
                   aes(label=variable)) +
  geom_hline(yintercept = 0.25, lty = 2, colour = "red")


grid.arrange(p2, p3, nrow = 1)
```

For both the two and three pointers the standardized bias for all covariates is below 0.25, indicating that are matched groups have similar enough covariate distributions to perform valid causal inference. The next step is to estimate a logistic regression model in the same form as was done above after the brute force matching, with whether the shot is made as the dependent variable, and the treatment group (isClutch) as well as all other covariates as the independent variables.

#### Logistic regression model results
```{r, echo=FALSE}
lr_threes_ps <- glm(shot_made_flag ~ isClutch + defender_distance + shot_distance + defender_distance + defender_height + dribbles + touch_time + action_type, 
               family = "binomial", data = matched_ps_threes)

lr_twos_ps <- glm(shot_made_flag ~ isClutch + defender_distance + shot_distance + defender_distance + defender_height + dribbles + touch_time + action_type, 
               family = "binomial", data = matched_ps_twos)

broom::tidy(lr_twos_ps) %>%
  mutate_if(is.numeric, round, 5) %>%
  head(7) %>% 
  kable(caption = "Model Parameters for two point shots",
        format = 'html') %>%
   row_spec(2, bold = TRUE) %>%
  kable_styling() 

```

```{r, echo=FALSE}
broom::tidy(lr_threes_ps) %>%
  mutate_if(is.numeric, round, 5) %>%
  head(7) %>% 
  kable(caption = "Model Parameters for three point shots",
        format = 'html') %>%
   row_spec(2, bold = TRUE) %>%
  kable_styling() 
```

The estimated effect of `Clutch` shots is similar in the propensity score matching to the brute force matching, with estimated of odds of a shot being $0.52$ (twos) or $0.67$ (threes) the odds of making an equivalent non-clutch shots. However, in this regression the estimates have more precision based on the much larger sample size, and are statistically significant at the $1\%$ significance level.
\newline


As expected based on the summary stats of the matched data, the estimate associated with the `isClutch` parameter is negative. Clutch shot attempts are estimated to have odds of success only 0.52 (twos) or 0.67 (threes) times the odds of success for their corresponding non-clutch shot attempts, holding all other covariates constant. Interestingly and a bit surprisingly to be honest, while there was a large difference in field goal percentage between clutch and non-clutch shot attempts even in the matched dataframe, the effect is not statistically significant at the $5\%$ significance level for both two point shots and three point shots, which is perhaps due to the reduced sample sizes. 

#### Discussions
Based on the analysis, it appears there may be a causal impact of `clutch` situations negatively impacting shooting percentage, even after accounting for the difficulty of the shot attempt with granular data on the shot situation and difficulty. However, there still may be unobservable characteristics of the shots which we have not captured in our covariates which may be systematically different between the clutch and non clutch attempts. While we have captured defender distance and height, it is not possible to estimate defender _effort_, which could plausibly be much different between an average shot, and an average clutch shot. Additionally, while we account for closest defender, it's possible many of these shots could have had double teams, where the second defender could still have an impact on shot make probability. It's hard to know if these sort of characteristics could be confounding the results, as they would require optical tracking data to capture, but it also seems that the mental stress and the body's subsequent physiological response in response to the `clutch` situations have a negative causal impact on shot making probabilities.  

<!-- - Could I have lumped both two point shots and three point shots together? Maybe, as that would have increased the sample sizes and precision of estimates, perhaps leading to statistical significance. I was only including only shots >= 10 feet, so they are all jump shots, and perhaps can all be compared.   -->
<!-- - I debated splitting the data even further and looking at situations when the shot taker is tied or leading the game versus shots when the shot taker is behind. The psychology and pressure associated with shooting to win (in a tied game), versus shooting not to lose (in a game down 2-3 points) may have heterogenous effects on the clutch shot impact on shooting percentage. -->

## References
#### Data
1. [NBA Savant](https://www.nbasavant.com/)
- Detailed shot logs, with granular data up until 2016
2. [Matthew Barlowe NBA Database](https://twitter.com/matthew_barlowe?lang=en)
- Has an Postgres database with 20 years of NBA PBP data as well as other stats. Posted online to the masses to message him for anyone to get access. Very grateful for this!

#### Causal Inference ideas
1. [Mike Lopez blog](https://statsbylopez.com/2017/01/04/matching-to-estimate-the-causal-effects-of-firing-an-nfl-coach/)
- This content was heavily inspired (especially the standardized bias stuff) from this blog post by Mike Lopez.

2. [Causal Kathy blog](https://causalkathy.com/)
- Multiple interesting causal inference based sports analysis blogs which helped guide the direction of this post.

3. [Matching methods review paper](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2943670/)
- Great overview of matching and propensity score methods, quite approachable as well for self-teaching.

